---
title: "Customer Personality Analysis"
author: "Junqian Zhang"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(lubridate)
library(nycflights13)
```

# Introduction

* Cluster analysis for user profile
* Build new variables which aggregate several variables to reduce the dimension

# Data Description

* 16 variables
* Personal Features
* Product Features
* Place Features

```{r, echo=FALSE, warning=FALSE}
data <- read.csv('marketing_data.csv')
data <- select(data, -c('ID', 'AcceptedCmp1', 'AcceptedCmp2', 'AcceptedCmp3', 'AcceptedCmp4', 'AcceptedCmp5', 'Response', 'Country', 'Complain', 'Recency', 'NumDealsPurchases', 'NumWebVisitsMonth'))
data$Income <- as.numeric(gsub('[$,]', '', data$Income))
str(data)
summary(data)
```

> *Personal variables*

* transfer Year_Birth to age
* transfer date of enrollment to Seniority
* combine Kidhome and Teenhome
* Marital Status: Alone, Divorced, Married, Single, Together, Widow ->  Living Status: Single, Couple
* Education: Basic, Graduation, Master, and PhD

```{r, echo=FALSE, warning=FALSE}
data$Age <- 2021-data$Year_Birth
data$Dt_Customer <- as.Date(data$Dt_Customer, '%m/%d/%y')
data$Seniority <- as.numeric(ymd('2021-09-01') - data$Dt_Customer)
data$Children <- data$Kidhome + data$Teenhome
data <- select(data, -c('Year_Birth', 'Dt_Customer', 'Teenhome', 'Kidhome'))
data <- data.frame(na.omit(data))
```

```{r,echo=FALSE, warning=FALSE}
data <- filter(data, Marital_Status!= "Absurd" & Marital_Status!="YOLO")
data$Living <- case_when(data$Marital_Status == 'Married' ~ 'couple', data$Marital_Status == 'Together' ~ 'couple', data$Marital_Status == 'Alone' ~ 'single', data$Marital_Status == 'Divorced' ~ 'single', data$Marital_Status == 'Together' ~ 'couple', data$Marital_Status == 'Widow'~'single', data$Marital_Status == 'Single'~'single')
data <- select(data, -"Marital_Status")
```

```{r,echo=FALSE, warning=FALSE}
data[data=='2n Cycle'] <- 'Master'
data[data=='Basic'] <- 'Graduation'
```

```{r, echo=FALSE, warning=FALSE}
data <- filter(data, MntWines <= 1285)
data <- filter(data, MntFruits <=172)
data <- filter(data, MntMeatProducts <= 914.78)
data <- filter(data, MntFishProducts <=226.78)
data <- filter(data, MntSweetProducts <=176)
data <- filter(data, MntGoldProds <=226.67)
data <- filter(data, NumWebPurchases <= 11)
data <- filter(data, NumCatalogPurchases <= 10.85)
data <- filter(data, NumStorePurchases <= 13)
data <- filter(data, Income <= 94462.32 )
data <- filter(data, Age <= 76 )
```

> *Product Variables*

* 6 variables: MntWines, MntFruits, MntMeatProducts, MntFishProducts, MntSweetProducts, MntGoldProds
* Replace 6 variables with 1 variable which can represent the product preference of the customers


##### Principle Component Analysis

```{r, echo=FALSE, warning=FALSE}
data_product <- select(data, c('MntWines', 'MntFruits', 'MntMeatProducts', 'MntFishProducts', 'MntSweetProducts', 'MntGoldProds'))
data <- select(data, -c('MntWines', 'MntFruits', 'MntMeatProducts', 'MntFishProducts', 'MntSweetProducts', 'MntGoldProds'))
summary(data_product)
```

```{r, echo=FALSE, warning=FALSE}
product_scale <- scale(data_product)
means<-colMeans(product_scale)
sds<-apply(product_scale,2,sd)
descr<-round(cbind(means,sds),2)

eigenval <- eigen(cor(product_scale))$values
eigenvec <- eigen(cor(product_scale))$vectors
#screediagram
eigen_dataframe <- data.frame(c(1:6), eigenval)
eigen_plot <- ggplot(data = eigen_dataframe, aes(x=as.factor(eigen_dataframe[,1]), y = eigen_dataframe[,2]))+
  geom_point(size=5)+
  geom_hline(yintercept = 1, color = "green", size= 2.5)+
  xlab("Components")+
  ylab("Eigevalue")
eigen_plot

#percentage of explained variance by each component 
eigen_power<-eigenval/6
eigenpower_dataframe<-data.frame(c(1:6),cumsum(eigen_power))
power_plot <- ggplot(data = eigenpower_dataframe, aes(x=as.factor(eigenpower_dataframe[,1]), y = eigenpower_dataframe[,2]))+
  geom_point(size=5)+
  geom_hline(yintercept = 0.8, color = "green", size= 2.5)+
  xlab("Components")+
  ylab("Variance explained")
power_plot

#Summary description of eigenvalue and eigenvectors in correspondence of variables
eigenvecdf <- data.frame(round(eigenvec[,1:3],3), row.names=colnames(data_product))
colnames(eigenvecdf)<-c("Comp1","Comp2","Comp3")
eigendf<-rbind(eigenvecdf,eigenval[1:3])
rownames(eigendf)[7]<-"EIGENVALUE"

comp<-round(cbind(eigenvec[,1]*sqrt(eigenval[1]),eigenvec[,2]*sqrt(eigenval[2]),eigenvec[,3]*sqrt(eigenval[3])),3)
rownames(comp)<-row.names(descr)
colnames(comp)<-c("Comp1","Comp2","Comp3")
comp_var<-rbind(comp,eigen_power[1:3])
rownames(comp_var)[6]<-"% of VAR explained"
communality<-comp[1:6,1]^2+comp[1:6,2]^2+comp[1:6,3]^2
comp_comu<-cbind(comp[1:6,],communality)
comp_comu
```
* Component 1: index of how much the customer dislike purchase
* Component 2: index of how much the customer dislike wine
* Component 3: index of how much the customer dislike gold


##### Clusters as new category values
```{r, echo=FALSE, warning=FALSE}
punteggi_3 <- product_scale%*%eigenvec[,1:3]
product.k <- data.frame("component1" = punteggi_3[,1], "component2" = punteggi_3[,2], "component3" = punteggi_3[,3])
library(factoextra)
fviz_nbclust(product.k, kmeans, method = "wss") +
    geom_vline(xintercept = 4, linetype = 2)+
  labs(subtitle = "Elbow method")
set.seed(42)
product_cluster <- kmeans(product.k,4)

product.k["cluster"] <- as.factor(product_cluster$cluster)
data["Product"] <- as.factor(product_cluster$cluster)

library(plotly)
fig <- plot_ly()
fig %>%
  add_trace(x = product.k[, 1], y = product.k[, 2], z = product.k[, 3], type = "scatter3d", mode = "markers",marker=list(opacity=0.6), color = product.k[, 4],size = 0.5) %>%
  add_trace(x = c(0,comp_comu[1,1]*10), y = c(0,comp_comu[1,2]*10), z = c(0,comp_comu[1,3]*10), type = "scatter3d", mode = "lines+text+markers",text= c("",rownames(eigendf)[1]),marker=list(symbol="diamond",size = 4), name = "lines", showlegend = FALSE, line = list(width = 10)) %>%
  add_trace(x = c(0,comp_comu[2,1]*10), y = c(0,comp_comu[2,2]*10), z = c(0,comp_comu[2,3]*10), type = "scatter3d", mode = "lines+text+markers", text= c("",rownames(eigendf)[2]),marker=list(symbol="diamond",size = 4), name = "lines", showlegend = FALSE, line = list(width = 10))%>%
  add_trace(x = c(0,comp_comu[5,1]*10), y = c(0,comp_comu[5,2]*10), z = c(0,comp_comu[3,3]*10), type = "scatter3d", mode = "lines+text+markers", text= c("",rownames(eigendf)[3]),marker=list(symbol="diamond",size = 4), name = "lines", showlegend = FALSE, line = list(width = 10))%>%
  add_trace(x = c(0,comp_comu[4,1]*10), y = c(0,comp_comu[4,2]*10), z = c(0,comp_comu[4,3]*10), type = "scatter3d", mode = "lines+text+markers", text= c("",rownames(eigendf)[4]),marker=list(symbol="diamond",size = 4), name = "lines", showlegend = FALSE, line = list(width = 10)) %>%
  add_trace(x = c(0,comp_comu[5,1]*10), y = c(0,comp_comu[5,2]*10), z = c(0,comp_comu[5,3]*10), type = "scatter3d", mode = "lines+text+markers", text= c("",rownames(eigendf)[5]),marker=list(symbol="diamond",size = 4), name = "lines", showlegend = FALSE, line = list(width = 10)) %>%
  add_trace(x = c(0,comp_comu[6,1]*10), y = c(0,comp_comu[6,2]*10), z = c(0,comp_comu[6,3]*10), type = "scatter3d", mode = "lines+text+markers", text= c("",rownames(eigendf)[6]),marker=list(symbol="diamond",size = 4), name = "lines", showlegend = FALSE, line = list(width = 10))

c1 <- ggplot(product.k, aes(x=cluster, y=component1, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "Component 1", x="Cluster")

c2 <- ggplot(product.k, aes(x=cluster, y=component2, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "Component 2", x="Cluster")

c3 <- ggplot(product.k, aes(x=cluster, y=component3, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "Component 3", x="Cluster")

figure <- ggarrange(c1, c2, c3, ncol=3, nrow=1)

figure




```


* Cluster 1: active shopper
* Cluster 2: wine lover
* Cluster 3: gold lover
* Cluster 4: inactive shopper


> *Variable Summary*

* 7 numeric variables
* 3 categorical variables

```{r, echo=FALSE, warning=FALSE}
data$Product <- as.factor(data$Product)
data$Education <- as.factor(data$Education)
data$Living <- as.factor(data$Living)
data$Product <- case_when(data$Product == 1 ~ 'active shopper', data$Product == 2 ~ 'wine lover', data$Product == 3 ~ 'gold lover', data$Product == 4 ~ 'inactive shopper')
data$Product <- as.factor(data$Product)
summary(data)
```

```{r, echo=FALSE, warning=FALSE}
b1 <- ggplot(data, aes(x = Education)) + ylab(NULL) +
  geom_bar(color="darkblue", fill="lightblue")
  
b2 <- ggplot(data, aes(x = Living)) + ylab(NULL) +
  geom_bar(color="darkblue", fill="lightblue")

b3<- ggplot(data, aes(x = Product)) + ylab(NULL) +
  geom_bar(color="darkblue", fill="lightblue")

figure <- ggarrange(b1,b2,b3,ncol=2, nrow=2)

figure
```

# Hierarchical Clustering

* Dissimilarity Measure: Gower Distance
* Linkage method: Ward's 
* Dunn's index for comparison

```{r, echo=FALSE, warning=FALSE}
library(StatMatch)
data_s <- cbind(data.frame(scale(data[,2:8])),select(data, c('Education', 'Product', 'Living')))
gower_dist <- gower.dist(data)
d <- as.dist(gower_dist)
```

```{r, echo=FALSE, warning=FALSE}
library(clValid)
hc_ward<- hclust(d,method="ward.D2")
x1 <- cutree(hc_ward, k=3)
x2 <- cutree(hc_ward, k=4)
d_3 <- dunn(distance = d, x1)
d_4 <- dunn(distance = d, x2)
dunn_df <- data.frame("3 clusters" = round(d_3,4), "4 clusters" = round(d_4,4))
colnames(dunn_df) <- c("3 clusters", "4 clusters")
rownames(dunn_df) <- "Dunn's Index"
dunn_df
```

##### Dendrogram
```{r, echo=FALSE, warning=FALSE}
plot(hc_ward, labels = rownames(data_s),hang=-5, cex = 0.6)
rect.hclust(hc_ward,k=4,border=c("orange","green","blue", "red"))
```

# Cluster Interpretation

> *Categorical variables*

```{r, echo=FALSE, warning=FALSE}
data_s$cluster <- as.factor(x2)
library(dplyr)
ed <- data_s %>% 
  group_by(cluster, Education) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = cluster, y = perc*100, fill = Education)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Education") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
               legend.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_fill_manual(values=c("#ff9999","#ccccff","#00ccff"))

living <- data_s %>% 
  group_by(cluster, Living) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = cluster, y = perc*100, fill = Living)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Living") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
               legend.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_fill_manual(values=c("#ff9999","#ccccff"))

prod <- data_s %>% 
  group_by(cluster, Product) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = cluster, y = perc*100, fill = Product)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Product") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
               legend.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_fill_manual(values=c("#ff9999","#ccccff","#00ccff", "#99ffff"))

figure <- ggarrange(ed, living, prod, ncol = 2, nrow = 2)
figure
```

> *Numeric variables*

```{r, echo=FALSE, warning=FALSE}
data$cluster <- as.factor(x2)
b1 <- ggplot(data, aes(x=cluster, y=Age, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "Age", x="Cluster")
b2 <- ggplot(data, aes(x=cluster, y=Income, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "Income", x="Cluster")
b3 <- ggplot(data, aes(x=cluster, y=NumWebPurchases, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "NumWebPurchases", x="Cluster")
b4 <- ggplot(data, aes(x=cluster, y=NumCatalogPurchases, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "NumCatalogPurchases", x="Cluster")
b5 <-ggplot(data, aes(x=cluster, y=NumStorePurchases, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "NumStorePurchases", x="Cluster")

b6 <- ggplot(data, aes(x=cluster, y=Seniority, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "Seniority", x="Cluster")
b7 <- ggplot(data, aes(x=cluster, y=Children, col=cluster)) +
  theme(legend.position="none") +
  geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(y = "Children", x="Cluster")
figure<- ggarrange(b1,b2,b3,b4,b5,b6,b7,ncol=3, nrow=3)
figure
```

* *Cluster 1*: customers in single who are inactive shoppers, and they have lower income but more children, and their frequency to the supermarket is relatively low; compared to catalog purchase and web purchase, store purchase is their preference
* *Cluster 2*: customers in couple who are inactive shoppers, they have children but lower income; same as customers in cluster 1, they do not have frequent purchase and prefer store purchase
* *Cluster 3*: customers in couple who buy in this company and go to the company frequently, they have longer membership and no children, and they are richer
* *Cluster 4*: customers in single who like shopping and shop frequently; same as cluster 3, they have longer membership and no children, also they have high income

# Features and Clusters

```{r, echo=FALSE, warning=FALSE}
library(rpart)
library(rpart.plot)
set.seed(42)
tree <- rpart(cluster~., data = data, method = 'class')
rpart.plot(tree)
```






